---
title: "**Data Analysis Blog**"
author: "TheDataAreClean"
date: "29th March, '20"
output: 
  html_document: 
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(echo = FALSE)

# Load packages
library(tidyverse)
library(formattable)
library(lubridate)
library(scales)
library(ggalt)
library(treemapify)
library(ggplotify)
```

### **Background**

Increase visibility into budgets data in India to improve civic participation.

* Citizens have limited access to data on government budgets in India.
* The state budget documents are not easily consumable with detailed information locked in closed formats (PDF, etc.).
* Lack of metadata hinders the process of easy search-ability for relevant fiscal data.
* Restricted public engagement with fiscal data leads constrained public participation with fiscal issues.

### **Objective**

Analysis of Himachal Pradesh state budgetary and district spending expenditure data.

* Comparative analysis for a sector of choice (Primary Education, Public Health or other).
* Identify some key fund flow issues and/or interesting data insights/stories.
* Use open-source softwares like Python, R, LibreOffice Calc or more, to keep the output shareable giving options to collate collaborative feedback.

### **Information**

#### **Related Blogs**

Existing work at CivicDataLab:

* [Tracking Flow of Public Funds through Open Data](https://medium.com/civicdatalab/tracking-flow-of-public-funds-through-open-data-1ace0e5b6163)
* [Roadmap for co-creating the Fiscal Data Explorer](https://medium.com/civicdatalab/roadmap-for-co-creating-the-fiscal-data-explorer-79818a53728f)
* [Design Research for Fiscal Data Explorer - A Work in Progress](https://medium.com/civicdatalab/design-research-for-fiscal-data-explorer-a-work-in-progress-43a07275b1c1)
* [Selecting schemes and sectors for Himachal Pradesh Fiscal Data Explorer - Process and Criteria](https://blog.openbudgetsindia.org/selecting-schemes-and-sectors-for-himachal-fiscal-data-explorer-process-and-criteria-de711f578166)

#### **Extended Reading list**

In-case one wishes to dig deeper, an [extended reading list](https://docs.google.com/document/d/1WYHy_7FLYI84XvXzcW277sqY2oA59zQFAcvbCVbJcR8/edit) explaining some of the key concepts of public finance in India and associated data.

### **Contents**

This report will cover the following areas:

1. **Data Health**
2. **Overall Analysis**
3. **Sector Analysis**
4. **Further Questions**
5. **Additional Ideas**
6. **Challenge**

### **Data Health Report**

#### **Files Recieved**

As a part of the assignment, 4 data files in `.csv` format were provided were provided and 2 metadata files in `.xlsx` format with descriptions of the data file fields.

```{r data-recieved}
# Data files for the assignment
dir("../data/raw/")

# Metadata files for the assignment
dir("../data/metadata/")
```

#### **Details**

**Datasets**
**_State Budget Expenditures_** & **_District Spending Expenditures_** for FY 2017-18 & FY 2018-19

**Information**
**_State Budget Expenditure_**: Information like Account Heads, Statement of Expenditure, Sanctions, Revisions;

**_District Spending Expenditures_**: Information like District, Treasury, DDO, Account Heads, Statement of Expenditure, Gross and Net Payments.

Additional details can be found in the [Data Health Report](https://www.google.com) in the GitHub repository.

### **Overall Analysis**

```{r read, cache = TRUE}
# State expenditure data
budget_exp_final <- read_csv("../data/processed/budget_exp_final.csv")

# Distirct spending data
district_spending_exp_final <- read_csv("../data/processed/district_spending_exp_final.csv")
```

```{r clean, cache = TRUE}
# Format the monetary value fields
# State Budget Expenditure
budget_exp_final$sanction <- accounting(budget_exp_final$sanction)
budget_exp_final$addition <- accounting(budget_exp_final$addition)
budget_exp_final$savings <- accounting(budget_exp_final$savings)
budget_exp_final$revised <- accounting(budget_exp_final$revised)

# District Spending  Expenditure
district_spending_exp_final$gross <- accounting(district_spending_exp_final$gross)
district_spending_exp_final$agded <- accounting(district_spending_exp_final$agded)
district_spending_exp_final$btded <- accounting(district_spending_exp_final$btded)
district_spending_exp_final$net_payment <- accounting(district_spending_exp_final$net_payment)
```

#### **State Budget Expenditure**

```{r overall}
# Quick Summary
state_budget_summary <- budget_exp_final %>%
                            group_by(year) %>%
                            summarise(sanction = sum(sanction),
                                      revised = sum(revised))

# Rename Fields
names(state_budget_summary) <- c("Fiscal Year", "Sanction", "Revised")

# Print formatted Table
formattable(state_budget_summary, list(
                                    Sanction = color_bar("orange"),
                                    Revised = color_bar("lightgreen")
                                    )
            )
```

* For the year 2018-19; the sanctioned budget was 1.59 Bn. & the revised budget was 1.51Bn.
* ~14.5% increase in the sanctioned state budget in the year 2018-19 over 2017-18.
* The revised state budget for 2018-17 was ~95% of the original sanctioned budget; and ~97.2% in 2017-18.

##### **Top Account Heads**

```{r top-account-heads}
# Account Head Wise spending
state_budget_demand <- budget_exp_final %>%
                            group_by(year, demand_desc) %>%
                            summarise(sanction = sum(sanction)) %>%
                            spread(year, sanction, fill = 0) %>%
                            arrange(desc(`2018-19`)) %>%
                            mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
                                   `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
                            mutate(`% of Total (2018-19)` = formattable::percent(`2018-19`/`Total Sanction (2018-19)`),
                                   `% of Total (2017-18)` = formattable::percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
                            select(demand_desc, `2018-19`, `% of Total (2018-19)`, `2017-18`, `% of Total (2017-18)`)

# Rename Fields
colnames(state_budget_demand)[1] <- "Demand"

# Plot bar chart for representation
# Absolutes
state_budget_demand %>%
        select(Demand, `2018-19`, `2017-18`) %>%
        head(10) %>%
        gather(Year, Amount, -Demand) %>%
        mutate(Demand = as_factor(Demand)) %>%
        ggplot(aes(x = Demand, y = Amount, fill = Year)) +
            geom_bar(stat = "identity", position = position_dodge()) +
            scale_x_discrete(label = function(x) stringr::str_trunc(x, 12)) +
            scale_y_continuous(labels = comma) +
            labs(title = "Top 10 Deparments by State Budget Expenditure", x = "Department", y = "Amount") +
            theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

# Percentage
state_budget_demand %>%
        select(Demand, `% of Total (2018-19)`, `% of Total (2017-18)`) %>%
        head(10) %>%
        gather(Year, Amount, -Demand) %>%
        mutate(Demand = as_factor(Demand)) %>%
        ggplot(aes(x = Demand, y = Amount, fill = Year)) +
            geom_bar(stat = "identity", position = position_dodge()) +
            scale_x_discrete(label = function(x) stringr::str_trunc(x, 12)) +
            scale_y_continuous(labels = percent) +
            labs(title = "Top 10 Deparments by State Budget Expenditure", x = "Department", y = "% Amount") +
            theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

* Finance was alloyed the highest amount of all account heads, 30.49% of total sanctioned expenditure for the year; followed by Education with 14.12% & Public Works – Roads Bridges & Buildings at 10.26%
* This trend has followed form 2017-18 in terms of priority for the state budget allocations.

##### **Voted vs Charged**

```{r voted-charged, warning = FALSE}
# Voted/Charged
budget_exp_final %>%
        group_by(year, voted_charged) %>%
        summarise(sanction = sum(sanction)) %>%
        spread(voted_charged, sanction) %>%
        mutate(percent_charged = formattable::percent(C/(V+C)),
               percent_voted = formattable::percent(V/(V+C))) %>%
        select(year, percent_voted, percent_charged) %>%
        gather(Bucket, Percentage, -year) %>%
        ggplot(aes(x = year, y = Percentage, fill = Bucket)) +
                geom_bar(stat = "identity", position = "stack") +
                scale_y_continuous(labels = percent) +
                scale_fill_discrete(name = "Type", labels = c("Charged", "Voted")) + 
                labs(title = "Voted vs Charged State Budget Expenditure", x = "Year", y = "%") +
                theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
                coord_flip()
```

~83% of all State Budget Expenditure is Voted for, remaining 17%  is charged.

##### **Plan vs Non-Plan**

```{r plan-non-plan, warning = FALSE}
# Plan/Non-Plan
budget_exp_final %>%
        group_by(year, plan_nonplan) %>%
        summarise(sanction = sum(sanction)) %>%
        spread(plan_nonplan, sanction) %>%
        mutate(percent_plan = formattable::percent(P/(N+P)),
               percent_nonplan = formattable::percent(N/(N+P))) %>%
        select(year, percent_plan, percent_nonplan) %>%
        gather(Bucket, Percentage, -year) %>%
        ggplot(aes(x = year, y = Percentage, fill = Bucket)) +
                geom_bar(stat = "identity", position = "stack") +
                scale_y_continuous(labels = percent) +
                scale_fill_discrete(name = "Type", labels = c("Non-Plan", "Plan")) + 
                labs(title = "Plan vs Non-Plan State Budget Expenditure", x = "Year", y = "%") +
                theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
                coord_flip()
                
```

Over 78% of all State Budget expenditure is Non-Plan, and remaining ~22% under Plan.

##### **YoY Comparison**

Changes in allotments in State Budget accross departments b/w 2018-19 and 2017-18.

```{r yoy-account-heads-inc}
# Overall Top 10 Investments
# Absolute change in account heads expenditure
state_budget_top_account_changes_absolute <- budget_exp_final %>%
                                                group_by(year, demand_desc) %>%
                                                summarise(sanction = sum(sanction)) %>%
                                                spread(year, sanction, fill = 0) %>%
                                                mutate(delta = `2018-19` - `2017-18`,
                                                      percent_delta = formattable::percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
                                                select(demand_desc, delta, percent_delta) %>%
                                                arrange(desc(delta)) %>%
                                                head(5)

# Rename Fields
names(state_budget_top_account_changes_absolute) <- c("Department", "Change", "% Change")

# Print formatted Table
formattable(state_budget_top_account_changes_absolute, list(
                                                            Change = color_bar("lightgreen"),
                                                            `% Change` = color_tile("paleturquoise", "turquoise")
                                                            )
            )

# % change in account heads expenditure
state_budget_top_account_changes_percent <- budget_exp_final %>%
                                                group_by(year, demand_desc) %>%
                                                summarise(sanction = sum(sanction)) %>%
                                                spread(year, sanction, fill = 0) %>%
                                                mutate(delta = `2018-19` - `2017-18`,
                                                      percent_delta = formattable::percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
                                                select(demand_desc, delta, percent_delta) %>%
                                                arrange(desc(percent_delta)) %>%
                                                head(5)

# Rename Fields
names(state_budget_top_account_changes_percent) <- c("Department", "Change", "% Change")

# Print formatted Table
formattable(state_budget_top_account_changes_percent, list(
                                                        Change = color_bar("lightgreen"),
                                                        `% Change` = color_tile("paleturquoise", "turquoise")
                                                        )
            )
```

* Finance, Education & Public Works has seen the most increase in absolute sanctions numbers in 2018-19 over 2017-18.
* Tourism and Civil Aviation (61.84%), Industries Mineral Supplies and Information Technology (52.77%) and Agriculture (34.04%) have seen highest sanctions in terms of % change.

```{r yoy-account-heads-dec}
# Overall Top 10 Disinvestments
# Absolute change in account heads expenditure
state_budget_bottom_account_changes_absolute <- budget_exp_final %>%
                                                    group_by(year, demand_desc) %>%
                                                    summarise(sanction = sum(sanction)) %>%
                                                    spread(year, sanction, fill = 0) %>%
                                                    mutate(delta = `2018-19` - `2017-18`,
                                                          percent_delta = formattable::percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
                                                    select(demand_desc, delta, percent_delta) %>%
                                                    arrange(delta) %>%
                                                    head(2)

# Rename Fields
names(state_budget_bottom_account_changes_absolute) <- c("Department", "Change", "% Change")

# Print formatted Table
formattable(state_budget_bottom_account_changes_absolute, list(
                                                            Change = color_bar("lightgreen"),
                                                            `% Change` = color_tile("paleturquoise", "turquoise")
                                                            )
            )
```
 
Election (37.58%), Labour Employment and Training (23.19%) have seen highest decrease sanctions in terms of % change.

##### **Sanction vs Revised**

Comparison of sanction vs revised 2018-19 State Budget Expenditure to understand departments with most changes.

```{r sanction-revised-positive-exp}
## 2018-19 positive absolute change
state_budget_postive_change_absolute <- budget_exp_final %>%
                                            filter(year == "2018-19") %>%
                                            group_by(demand_desc) %>%
                                            summarise(sanction = sum(sanction),
                                                      revised = sum(revised)) %>%
                                            mutate(revision_delta = (revised - sanction),
                                                   percent_delta =  formattable::percent((revised - sanction)/sanction)) %>%
                                            arrange(desc(revision_delta)) %>%
                                            select(demand_desc, revision_delta, percent_delta) %>%
                                            head(5)
    
# Rename Fields
names(state_budget_postive_change_absolute) <- c("Department", "Change", "% Change")

# Print formatted Table
formattable(state_budget_postive_change_absolute, list(
                                                    Change = color_bar("gold"),
                                                    `% Change` = color_tile("violet", "darkviolet")
                                                    )
            )

## 2018-19 positive percent change
state_budget_postive_change_percent <- budget_exp_final %>%
                                            filter(year == "2018-19") %>%
                                            group_by(demand_desc) %>%
                                            summarise(sanction = sum(sanction),
                                                      revised = sum(revised)) %>%
                                            mutate(revision_delta = (revised - sanction),
                                                   percent_delta =  formattable::percent((revised - sanction)/sanction)) %>%
                                            arrange(desc(percent_delta)) %>%
                                            select(demand_desc, revision_delta, percent_delta) %>%
                                            head(5)

    
# Rename Fields
names(state_budget_postive_change_percent) <- c("Department","Change", "% Change")

# Print formatted Table
formattable(state_budget_postive_change_percent, list(
                                                    Change = color_bar("gold"),
                                                    `% Change` = color_tile("violet", "darkviolet")
                                                    )
            )
```

* Land Revenue and District Administration & Finance saw one of the highest increase from sanctions to revisions.
* Both financial years saw over huge increases in election expenditure from sanctions to revised 74.37% in 2018-19.

```{r sanction-revised-negative-exp}
## 2018-19 negative absolute change
state_budget_negative_change_absolute <- budget_exp_final %>%
                                            filter(year == "2018-19") %>%
                                            group_by(demand_desc) %>%
                                            summarise(sanction = sum(sanction),
                                                      revised = sum(revised)) %>%
                                            mutate(revision_delta = (revised - sanction),
                                                   percent_delta =  formattable::percent((revised - sanction)/sanction)) %>%
                                            arrange(revision_delta) %>%
                                            select(demand_desc, revision_delta, percent_delta) %>%
                                            head(5)
    
# Rename Fields
names(state_budget_negative_change_absolute) <- c("Department", "Change", "% Change")

# Print formatted Table
formattable(state_budget_negative_change_absolute, list(
                                                    Change = color_bar("gold"),
                                                    `% Change` = color_tile("violet", "darkviolet")
                                                    )
            )

## 2018-19 positive percent change
state_budget_negative_change_percent <- budget_exp_final %>%
                                            filter(year == "2018-19") %>%
                                            group_by(demand_desc) %>%
                                            summarise(sanction = sum(sanction),
                                                      revised = sum(revised)) %>%
                                            mutate(revision_delta = (revised - sanction),
                                                   percent_delta =  formattable::percent((revised - sanction)/sanction)) %>%
                                            arrange(percent_delta) %>%
                                            select(demand_desc, revision_delta, percent_delta) %>%
                                            head(5)

    
# Rename Fields
names(state_budget_negative_change_percent) <- c("Department", "Change", "% Change")

# Print formatted Table
formattable(state_budget_negative_change_percent, list(
                                                    Change = color_bar("gold"),
                                                    `% Change` = color_tile("violet", "darkviolet")
                                                    )
            )
```

* Education saw the highest decrease from sanction to revised in both years with 34 Mn, ~15.22% in 2018-19.
* Scheduled Castes Sub Plan, Rural Development, Tribal Development all saw decreases from sanction to revised.

#### **District Spending Expenditure**

##### **Overall**

```{r district-overall}
# Quick Summary
district_spending_summary <- district_spending_exp_final %>%
                                group_by(year) %>%
                                summarise(gross = sum(gross),
                                net_payment = sum(net_payment))

# Rename Fields
names(district_spending_summary) <- c("Fiscal Year", "Gross", "Net Payment")

# Print formatted Table
formattable(district_spending_summary, list(
                                            Gross = color_bar("orange"),
                                            `Net Payment` = color_bar("lightgreen")
                                            )
            )
```

The net payments amount for Himachal Pradesh in 2018-19 was 330.9 Bn.; a 12.7% increase over the previous years District Spending.

##### **Spending Over Time**

```{r district-over-time}
# Trends for budget allocation over time
district_spending_exp_final %>%
    mutate(month = floor_date(transaction_date, unit = "month")) %>%
    group_by(month) %>%
    summarise(gross = sum(gross),
              net_payment = sum(net_payment)) %>%
    gather(type, amount, -month) %>%
    ggplot(aes(x = month, y = amount, color = type)) +
    geom_line() + geom_point() +
    scale_y_continuous(labels = comma) +
    scale_color_discrete(name = "Cylinders", labels = c("Gross", "Net Payment")) +
    labs(title = "District Spending Expenditure Over Time", x = "Time", y = "Amount") +
    theme_minimal()
```

Majority of district level spending expenditure happened during the month of March (EoY), a trend that can traced through both fiscal years.

##### **Top Districts & Treasuries**

```{r distirct-treasuries}
# District Wise spending
district_spending_exp_final %>%
    filter(year == "2018-19") %>%
    group_by(district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment)) %>%
    mutate(net_payment_total = sum(net_payment)) %>%
    mutate(net_payment_percent = formattable::percent(net_payment/net_payment_total)) %>%
    select(district, net_payment_percent) %>%
    mutate(district = as_factor(district)) %>%
    ggplot(aes(x = district, y = net_payment_percent, label = net_payment_percent)) + 
        geom_point(col = "blue", size = 3) + geom_text(vjust = -0.3) +
        geom_segment(aes(x = district, xend = district, y = 0, yend = net_payment_percent)) + 
    scale_y_continuous(labels = percent) +
    labs(title="District Wise Share of Spending Expenditure", x = "District", y = "%") +
    theme_minimal() + theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

# Treasury Wise spending
district_spending_exp_final %>%
    filter(year == "2018-19") %>%
    group_by(treasury) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment)) %>%
    mutate(net_payment_total = sum(net_payment)) %>%
    mutate(net_payment_percent = formattable::percent(net_payment/net_payment_total)) %>%
    select(treasury, net_payment_percent) %>%
    head(10) %>%
    mutate(treasury = as_factor(treasury)) %>%
    ggplot(aes(x = reorder(treasury, desc(treasury)), 
               y = net_payment_percent, 
               label = net_payment_percent)) + 
        geom_point(col = "tomato2", size = 3) + geom_text(hjust = -0.2) +
        geom_segment(aes(x = treasury, xend = treasury, 
                         y = min(net_payment_percent), yend = max(net_payment_percent)), 
                     linetype = "dashed", size = 0.1) + 
    scale_y_continuous(labels = percent) +
    labs(title="Top 10 Treasuries for Spending Expenditure", x = "Treasury", y = "%") +
    theme_minimal() + theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
    coord_flip()
```

##### **Top Account Heads**

```{r district-account-heads}
# Account Head wise spending
district_spending_account_yoy <- district_spending_exp_final %>%
                                    group_by(year, demand_desc) %>%
                                    summarise(net_payment = sum(net_payment)) %>%
                                    spread(year, net_payment, fill = 0) %>%
                                    arrange(desc(`2018-19`)) %>%
                                    mutate(net_payment_18_19 = sum(`2018-19`),
                                           net_payment_17_18 = sum(`2017-18`)) %>%
                                    mutate(net_payment_18_19_percent = 
                                               formattable::percent(`2018-19`/net_payment_18_19),
                                           net_payment_17_18_percent = 
                                               formattable::percent(`2017-18`/net_payment_17_18)) %>%
                                    select(demand_desc, net_payment_18_19_percent, net_payment_17_18_percent) %>%
                                    head(8)

# Plot the major account heds
# Prep Data
left_label <- paste(stringr::str_trunc(district_spending_account_yoy$demand_desc, 12),
                    district_spending_account_yoy$net_payment_17_18_percent,sep = ", ")
right_label <- paste(stringr::str_trunc(district_spending_account_yoy$demand_desc, 12),
                     district_spending_account_yoy$net_payment_18_19_percent,sep=", ")
district_spending_account_yoy$class <- ifelse((
    district_spending_account_yoy$net_payment_18_19_percent - 
        district_spending_account_yoy$net_payment_17_18_percent) < 0, "red", "green")

# Plot
p <- ggplot(district_spending_account_yoy) + 
        geom_segment(aes(x = 1, xend = 2, 
                         y = net_payment_17_18_percent, yend = net_payment_18_19_percent, 
                         col = class), size=.75, show.legend = F) + 
                    geom_vline(xintercept = 1, linetype = "dashed", size = .1) + 
                    geom_vline(xintercept = 2, linetype = "dashed", size = .1) +
                    scale_color_manual(labels = c("Up", "Down"), 
                                       values = c("green" = "#00ba38", "red" = "#f8766d")) +
                    labs(x = "", y = "", title = "Top Department Spending over Fiscal Years") +
                    xlim(.5, 2.5) +
                    ylim(0,(1.1*(max(district_spending_account_yoy$net_payment_17_18_percent,
                                     district_spending_account_yoy$net_payment_18_19_percent)))) +
                    theme_minimal()

# Add texts
p <- p + geom_text(label = left_label, check_overlap = TRUE,
                   y = district_spending_account_yoy$net_payment_17_18_percent,
                   x = rep(1, NROW(district_spending_account_yoy)), 
                   hjust = 1.1, size = 3.5)

p <- p + geom_text(label = right_label,  check_overlap = TRUE,
                   y = district_spending_account_yoy$net_payment_18_19_percent,
                   x = rep(2, NROW(district_spending_account_yoy)), 
                   hjust = -0.1, size = 3.5)

p <- p + geom_text(label="2017-18", 
                   x = 1, 
                   y = 1.1*(max(district_spending_account_yoy$net_payment_17_18_percent, 
                                district_spending_account_yoy$net_payment_18_19_percent)), 
                   hjust = 1.2, size = 5)  # title

p <- p + geom_text(label="2018-19", 
                   x = 2, 
                   y = 1.1*(max(district_spending_account_yoy$net_payment_17_18_percent, 
                                district_spending_account_yoy$net_payment_18_19_percent)), 
                   hjust = -0.1, size = 5)  # title

# Minify theme
p + theme(panel.background = element_blank(), 
           panel.grid = element_blank(),
           axis.ticks = element_blank(),
           axis.text.x = element_blank(),
           panel.border = element_blank(),
           plot.margin = unit(c(1,2,1,2), "cm"))
```

Finance (16.04%), Education (11.49%) and Health and Family Welfare (4.59%) have the highest amount of net payments across both financial years.

```{r na-account-heads}
# Demand Description is NA
district_spending_exp_final %>%
    filter(is.na(demand_desc)) %>%
    group_by(year, district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(net_payment_18_19 = sum(`2018-19`),
           net_payment_17_18 = sum(`2017-18`)) %>%
    mutate(net_payment_18_19_percent = 
               formattable::percent(`2018-19`/net_payment_18_19),
           net_payment_17_18_percent = 
               formattable::percent(`2017-18`/net_payment_17_18)) %>%
    select(district, net_payment_18_19_percent, net_payment_17_18_percent) %>%
    gather(year, percent, -district) %>%
    mutate(district = as_factor(district)) %>%
    ggplot(aes(x = district, y = percent, fill = year)) +
            geom_bar(stat = "identity", position = position_dodge()) +
            scale_y_continuous(labels = percent) +
            scale_fill_discrete(name = "Fiscal Year", labels = c("2017-18", "2018-19")) + 
            labs(title = "Department NA across Districts", x = "Department", y = "% Amount") +
            theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

26.39% of net payments have no Department info against them, when looking at the district data they are spread across all with Shimla at top with 24.06% of the NA spend and Dharamsala at 14.91%.

##### **Change Over Years**

Top 10 departments with change is spending over the 2 fiscal years.

```{r district-account-heads-change}
# Abslolute Positive Change in Net Payments
account_head_change_yoy <- district_spending_exp_final %>%
      filter(!is.na(demand_desc)) %>%
      group_by(year, demand_desc) %>%
      summarise(net_payment = sum(net_payment)) %>%
      spread(year, net_payment, fill = 0) %>%
      mutate(delta = `2018-19` - `2017-18`,
             percent_delta = formattable::percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
      select(demand_desc, `2018-19`, `2017-18`, delta, percent_delta) %>%
      arrange(desc(`2018-19`)) %>%
      head(10)

# Ordering of values
account_head_change_yoy$demand_desc <- 
              factor(stringr::str_trunc(account_head_change_yoy$demand_desc, 12),
                     levels = rev(as.character(stringr::str_trunc(
                       account_head_change_yoy$demand_desc, 12))
                       )
                     )

# Plotting of values
ggplot(account_head_change_yoy, 
         aes(x = `2017-18`, xend = `2018-19`, 
             y = demand_desc, 
             group = demand_desc)) + 
      geom_dumbbell(color = "#a3c4dc", 
                    size = 1, 
                    colour_xend = "#0e668b") + 
      scale_x_continuous(label = comma) + 
      labs(x = "Spending Expenditure", 
           y = "Department", 
           title = "Top 10 Departments: Change in Sending over Fiscal Years",
           subtitle = "2017-18 vs 2018-19") +
      theme_minimal()
```

* Power Development, Land Revenue and District Administration, Rural Development had the highest absolute increase in net payments YoY.
* Power Development (160.48%), Urban Development Town and Country Planning and Housing (77.97%) highest % increase in net payments YoY.

### **Sector Analysis**

Analysis of Education Department entries in the budgetary datasets.

#### **State Budget Expenditure**

##### **Overall**
```{r ed-overall}
# Quick Summary
education_state_budget_overall <- budget_exp_final %>%
          filter(demand_desc == "EDUCATION") %>%
          group_by(year) %>%
          summarise(sanction = sum(sanction),
                    revised = sum(revised))

# Rename Fields
names(education_state_budget_overall) <- c("Fiscal Year", "Sanction", "Revised")

# Print formatted Table
formattable(education_state_budget_overall, list(
                                                Sanction = color_bar("orange"),
                                                Revised = color_bar("lightgreen")
                                                )
            )
```

In 2018-19, the revised education budget was down to 85% of originally sanctioned; compared to 95% in 2017-18.

##### **Account Heads**

```{r ed-account-heads, warning = FALSE}
# Major/Sub-Major Expenditure 
ed_major_state_expenditure <- budget_exp_final %>%
          filter(demand_desc == "EDUCATION" & year == "2018-19") %>%
          group_by(major_desc, sub_major_desc) %>%
          summarise(sanction = sum(sanction)) %>%
          mutate(total_sanction = sum(sanction)) %>%
          mutate(percent_sanction = formattable::percent(sanction/total_sanction)) %>%
          arrange(desc(sanction))

# Plot the treemap
ggplot(ed_major_state_expenditure, 
       aes(area = sanction, fill = percent_sanction, 
           label = sub_major_desc, subgroup = major_desc)) +
  geom_treemap() +
  geom_treemap_subgroup_border() +
  scale_fill_continuous(name = "Percentage") +
  geom_treemap_subgroup_text(place = "bottom", grow = T, alpha = 0.5, colour =
                             "black", fontface = "italic", min.size = 0) +
  geom_treemap_text(colour = "white", place = "topleft", reflow = T)
```

* ~98% of all state education expenditure was towards General Education function.
* The top sub-sectors were Elementary Education (53.76%), Secondary Education (38.54%) & University and Higher Education (7.24%).

```{r ed-account-heads-minor, warning = FALSE}
# Major/Sub-Major Expenditure 
ed_major_state_expenditure <- budget_exp_final %>%
          filter(demand_desc == "EDUCATION" & 
                   year == "2018-19" &
                   sub_major_desc == "ELEMENTARY EDUCATION") %>%
          group_by(minor_desc, sub_minor_desc) %>%
          summarise(sanction = sum(sanction)) %>%
          mutate(total_sanction = sum(sanction)) %>%
          mutate(percent_sanction = formattable::percent(sanction/total_sanction)) %>%
          arrange(desc(sanction))

# Plot the treemap
ggplot(ed_major_state_expenditure, 
       aes(area = sanction, fill = percent_sanction, 
           label = sub_minor_desc, subgroup = minor_desc)) +
  geom_treemap() +
  geom_treemap_subgroup_border() +
  scale_fill_continuous(name = "Percentage") +
  geom_treemap_subgroup_text(place = "bottom", grow = T, alpha = 0.5, colour =
                             "black", fontface = "italic", min.size = 0) +
  geom_treemap_text(colour = "white", place = "topleft", reflow = T)
```

```{r ed-account-heads-minor-2, warning = FALSE}
# Major/Sub-Major Expenditure 
ed_major_state_expenditure <- budget_exp_final %>%
          filter(demand_desc == "EDUCATION" & 
                   year == "2018-19" &
                   sub_major_desc == "SECONDARY EDUCATION") %>%
          group_by(minor_desc, sub_minor_desc) %>%
          summarise(sanction = sum(sanction)) %>%
          mutate(total_sanction = sum(sanction)) %>%
          mutate(percent_sanction = formattable::percent(sanction/total_sanction)) %>%
          arrange(desc(sanction))

# Plot the treemap
ggplot(ed_major_state_expenditure, 
       aes(area = sanction, fill = percent_sanction, 
           label = sub_minor_desc, subgroup = minor_desc)) +
  geom_treemap() +
  geom_treemap_subgroup_border() +
  scale_fill_continuous(name = "Percentage") +
  geom_treemap_subgroup_text(place = "bottom", grow = T, alpha = 0.5, colour =
                             "black", fontface = "italic", min.size = 0) +
  geom_treemap_text(colour = "white", place = "topleft", reflow = T)
```

Under Elementary Education, 85% of the expenditure was towards Govt. Primary Schools & for Secondary Education, 98% of expenditure was sanctioned towards Govt. Secondary Schools.

##### **Voted vs Charged**

```{r ed-voted-charged, warning = FALSE}
# Voted/Charged
budget_exp_final %>%
        filter(demand_desc == "EDUCATION" & year == "2018-19") %>%
        group_by(year, voted_charged) %>%
        summarise(sanction = sum(sanction)) %>%
        spread(voted_charged, sanction) %>%
        mutate(percent_charged = formattable::percent(C/(V+C)),
               percent_voted = formattable::percent(V/(V+C))) %>%
        select(year, percent_voted, percent_charged) %>%
        gather(Bucket, Percentage, -year) %>%
        ggplot(aes(x = year, y = Percentage, fill = Bucket)) +
                geom_bar(stat = "identity", position = "stack") +
                scale_y_continuous(labels = percent) +
                scale_fill_discrete(name = "Type", labels = c("Charged", "Voted")) + 
                labs(title = "Voted vs Charged State Budget Expenditure", x = "Year", y = "%") +
                theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
                coord_flip()
```

All the Himachal Pradesh state expenditure towards education was voted for.

##### **Plan vs Non-Plan**

```{r ed-plan-non-plan, warning = FALSE}
# Plan/Non-Plan
budget_exp_final %>%
        filter(demand_desc == "EDUCATION" ) %>%
        group_by(year, plan_nonplan) %>%
        summarise(sanction = sum(sanction)) %>%
        spread(plan_nonplan, sanction) %>%
        mutate(percent_plan = formattable::percent(P/(N+P)),
               percent_nonplan = formattable::percent(N/(N+P))) %>%
        select(year, percent_plan, percent_nonplan) %>%
        gather(Bucket, Percentage, -year) %>%
        ggplot(aes(x = year, y = Percentage, fill = Bucket)) +
                geom_bar(stat = "identity", position = "stack") +
                scale_y_continuous(labels = percent) +
                scale_fill_discrete(name = "Type", labels = c("Non-Plan", "Plan")) + 
                labs(title = "Plan vs Non-Plan State Budget Expenditure", x = "Year", y = "%") +
                theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
                coord_flip()
                
```

~16% of the state expenditure towards Education was planned and the remaining non-planned.

#### **District Spending**

##### **Overall**

```{r ed-district-overall}
# Quick Summary
education_district_spending_overall <- district_spending_exp_final %>%
            filter(demand_desc == "EDUCATION") %>%
            group_by(year) %>%
            summarise(gross = sum(gross),
                      net_payment = sum(net_payment))

# Rename Fields
names(education_district_spending_overall) <- c("Fiscal Year", "Gross", "Net Payments")

# Print formatted Table
formattable(education_district_spending_overall, list(
                                                    Gross = color_bar("orange"),
                                                    `Net Payments` = color_bar("lightgreen")
                                                    )
            )
```

* ~4% increase in payments in 2018-19 over 2017-18
* Net payments comprise of ~72% of gross in FY 2018-19; and ~71% in 2017-18

##### **Spending Over Time**

```{r ed-district-over-time}
# Trends for budget allocation over time
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    mutate(month = floor_date(transaction_date, unit = "month")) %>%
    group_by(month) %>%
    summarise(gross = sum(gross),
              net_payment = sum(net_payment)) %>%
    gather(type, amount, -month) %>%
    ggplot(aes(x = month, y = amount, color = type)) +
    geom_line() + geom_point() +
    scale_y_continuous(labels = comma) +
    scale_color_discrete(name = "Cylinders", labels = c("Gross", "Net Payment")) +
    labs(title = "District Spending Expenditure Over Time", x = "Time", y = "Amount") +
    theme_minimal()
```

Majority of district level Education spending expenditure happened during the month of March (EoY), a trend that can traced through both fiscal years.

##### **Top Districts**

```{r ed-distirct-treasuries}
# District wise education spending
district_ed <- district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19") %>%
    group_by(district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment)) %>%
    mutate(net_payment_total = sum(net_payment)) %>%
    mutate(net_payment_percent = formattable::percent(net_payment/net_payment_total)) %>%
    select(district, net_payment, net_payment_percent)

# Education as % of every district's spending
percent_ed <- district_spending_exp_final %>%
    filter(year == "2018-19") %>%
    mutate(ed_flag = if_else(is.na(demand_desc) | demand_desc != "EDUCATION", 
                             "OTHER", "EDUCATION")) %>%
    group_by(ed_flag, district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(ed_flag, net_payment, fill = 0) %>%
    mutate(percent_ed = formattable::percent(EDUCATION/(EDUCATION + OTHER))) %>%
    arrange(desc(EDUCATION)) %>%
    select(district, percent_ed)

# Join the 2 tables
district_wise_ed <- district_ed %>%
                        left_join(percent_ed, by = "district")

# Rename fields
names(district_wise_ed) <- c("District", "Net Payments", "Ed. Share Overall", "Ed. Share District")

# Print formatted table
formattable(district_wise_ed, list(
                                  `Net Payments` = color_bar("lightgreen"),
                                  `Ed. Share Overall` = color_tile("white", "lightblue"), 
                                  `Ed. Share District` = color_tile("white", "lightpink")
                                  )
            )
```

* Shimla Capital Treasury and Dharamsala have respectively 17.78% and 16.37% of all education spending expenditure.
* 19.91% spending in Kullu and 19.34% spending in Bilaspur is towards education.
* Treasuries like Kamrau, Kupvi and Bhalie have over 70% of their budget spending in education.

##### **Statement of Expenditure**

```{r ed-district-soe}
# Education Overall
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, soe_description) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = formattable::percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = formattable::percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(soe_description, `% of net_payment (2018-19)`, `% of net_payment (2017-18)`) %>%
    head(10) %>%
    gather(year, percent, -soe_description) %>%
    mutate(soe_description = as_factor(soe_description)) %>%
    ggplot(aes(x = soe_description, y = percent, fill = year)) +
            geom_bar(stat = "identity", position = position_dodge()) +
            scale_x_discrete(label = function(x) stringr::str_trunc(x, 12)) +
            scale_y_continuous(labels = percent) +
            labs(title = "Top 10 SOEs by District Spending Expenditure", x = "SOE", y = "% Amount") +
            scale_fill_discrete(name = "Fiscal Year", labels = c("2017-18", "2018-19")) +
            theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

* ~78% of all education spending was towards salaries.
* 9.6% of spending is towards GIA GENERAL (Salary).

##### **Schemes**

```{r ed-schemes}
# Education Overall
ed_schemes <- district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
            year == "2018-19" &
            (str_detect(sub_minor_desc, pattern = c("YOJNA")) |
               str_detect(sub_minor_desc, pattern = c("MID DAY")) |
               str_detect(sub_minor_desc, pattern = c("ABHYAN")) |
               str_detect(sub_minor_desc, pattern = c("ABHIYAN")))) %>%
    group_by(sub_minor_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment))

# Rename fields
names(ed_schemes) <- c("Scheme", "Net Payments")

# Print formatted table
formattable(ed_schemes, list(
                            `Net Payments` = color_bar("lightgreen")
                            )
            )
```

### **Further Questions**

* In-depth analysis of function and further granularity to identify key spending trends by the education department.
* DDO level analysis to understand gaps in on-ground dissemination of funds to better identify areas of opportunity.
* Comparison of public finance data with further state level population, demographics information to analyse against ground reality.

### **Additional Ideas**

* Develop better understanding from other initiatives exploring public finance data for improvement of child education and open budget explorers to identify key intentions and move learn from the same.
* Understanding the different personas and user journeys on OpenBudgetsIndia by different kind of users & use a data driven approach to customise the platform experience to the user and familiarity with the information at hand.

### **Challenges**

**Sectorial**

* The assignment left me with a lot more questions to answer with my limited knowledge of the financial sector.
* The documentation was helpful in understanding the basic of public finance data; learning from an industry expert would benefit in finding better insights.

**Technical**

* The entire analysis was done using R and the mainly the tidyverse packages.
* Lack of availability of a dashboarding tool created a challenge to replicate some ideas and create narrative in-depth data stories.

### **Session Info**

```{r session}
# Information about the machine and enviroment
sessionInfo()
```
