---
title: "**Data Health Report**"
author: "TheDataAreClean"
date: "23rd March, '20"
output: 
  html_document: 
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 6
---

### **Objective**

Analysing the health and integrity of the data files recieved or the Public Finance Researcher position at CivicDataLab.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

# Load packages
library(tidyverse)
library(readxl)
library(formattable)
```

### **Datasets**

As a part of the assignment, 4 data files in `.csv` format were provided were provided and 2 metadata files in `.xlsx` format with descriptions of the data file fields.

```{r data}
# Data files for the assignment
dir("../data/raw/")

# Metadata files for the assignment
dir("../data/metadata/")
```

#### **Read**

Read the individual state expenditure and district spending data files into the enviroment for analysis.

```{r read, cache = TRUE}
# Read the 2018-19 & 2017-18 state wise budget data
budget_exp_2018_19 <- read_csv("../data/raw/budget_expenditure_fy_2018_19.csv")
budget_exp_2017_18 <- read_csv("../data/raw/budget_expenditure_fy_2017_18.csv")

# Read the 2018-19 & 2017-18 district wise spending data
district_spending_exp_2018_19 <- read_csv("../data/raw/district_spending_expenditure_fy_2018_19.csv")
district_spending_exp_2017_18 <- read_csv("../data/raw/district_spending_expenditure_fy_2017_18.csv")
```

The metadata descriptions for the files.

```{r metadata}
## State Expenditure
# Read the metadata
state_expenditure_metadata <- read_excel("../data/metadata/Metadata Himachal Budget Allocation.xlsx")

# Review the fields
state_expenditure_metadata

## District Spending
# Read the metadata
district_spending_metadata <- read_excel("../data/metadata/Metadata Himachal Spending Expendiure.xlsx")

# Review the fields
district_spending_metadata
```

#### **Review**

Review the composition of individual state expenditure and district spending datasets.

```{r review, cache = TRUE}
glimpse(budget_exp_2018_19)
glimpse(budget_exp_2017_18)
glimpse(district_spending_exp_2018_19)
glimpse(district_spending_exp_2017_18)
```

#### **Clean**

Clean & normalize the dataset using learnings from the metadata & review to prepare the final datasets for analysis.

```{r clean, cache = TRUE}
## State expenditure data
# Add the year field signifier
budget_exp_2018_19$year <- "2018-19"
budget_exp_2017_18$year <- "2017-18"

# Append the 2 financial year datasets as for analysis
budget_exp_final <- bind_rows(budget_exp_2018_19, budget_exp_2017_18)

# Rename the fields
names(budget_exp_final)[14:20] <- c("soe", "soe_description", "transaction_date", "sanction", "addition", "savings", "revised")

# Normalize text description fields
budget_exp_final$demand_desc <- toupper(trimws(budget_exp_final$demand_desc))
budget_exp_final$major_desc <- toupper(trimws(budget_exp_final$major_desc))
budget_exp_final$sub_major_desc <- toupper(trimws(budget_exp_final$sub_major_desc))
budget_exp_final$minor_desc <- toupper(trimws(budget_exp_final$minor_desc))
budget_exp_final$sub_minor_desc <- toupper(trimws(budget_exp_final$sub_minor_desc))

## District spending data
# Add the year field signifier
district_spending_exp_2018_19$year <- "2018-19"
district_spending_exp_2017_18$year <- "2017-18"

# Append the 2 financial year datasets as for analysis
district_spending_exp_final <- bind_rows(district_spending_exp_2018_19, district_spending_exp_2017_18)

# Rename the fields
names(district_spending_exp_final)[1:5] <- c("district", "treasury_code", "treasury", "ddo_code", "ddo_desc")
names(district_spending_exp_final)[19:27] <- c("soe", "soe_description", "transaction_date", "voucher", "bills", "gross", "agded", "btded", "net_payment")

# Normalize text description fields
district_spending_exp_final$ddo_desc <- toupper(trimws(district_spending_exp_final$ddo_desc))
district_spending_exp_final$demand_desc <- toupper(trimws(district_spending_exp_final$demand_desc))
district_spending_exp_final$major_desc <- toupper(trimws(district_spending_exp_final$major_desc))
district_spending_exp_final$sub_major_desc <- toupper(trimws(district_spending_exp_final$sub_major_desc))
district_spending_exp_final$minor_desc <- toupper(trimws(district_spending_exp_final$minor_desc))
district_spending_exp_final$sub_minor_desc <- toupper(trimws(district_spending_exp_final$sub_minor_desc))
```

### **Findings**

#### **Data**

* Granular transactional level state expenditure and district spending information.
* Information avialble at Account Heads and Statement of Expenditure granularity.
* Details available on Voted / Charged and Plan / Non-Plan expenditure
* Monetary information available broken down by sactioned, added and saved amount.
* Text description fields normalized for consumption throughout the analysis.

#### **Ideas**

**_Overall Analysis_**

* YoY analysis of monetary trends; sactions, additions, savings, revisions.
* Sector wise analysis of finds deployed and trends observed.
* Gap identification in the expendiure vs spendings datasets.
* Unavailable information across files for stated expenditure.

**_Sector Analysis_**

* Education sector analysis on funds availability and dissemination.
* Trends of spending over time and by usage of funds.
* Geography level analysis over where the funds are being disseminated.

#### **Next Steps**

Export cleaned datasets for further analysis of funds basis the ideas stated above.

```{r export, eval = FALSE}
# State expenditure combined & cleaned dataset
write_csv(budget_exp_final, "../data/processed/budget_exp_final.csv")

# District expenditure combined & cleaned dataset
write_csv(district_spending_exp_final, "../data/processed/district_spending_exp_final.csv")
```

### **Session Info**

```{r}
# Information about the machine and enviroment
sessionInfo()
```

