---
title: '**Deep Dive Report**'
author: "TheDataAreClean"
date: "7th April, '20"
output:
  html_document:
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

# Load packages
library(tidyverse)
library(formattable)
library(lubridate)
library(scales)
library(ggalt)
library(treemapify)
library(ggplotify)
```

### **Objective**

Follow-up on initial assignment with a deep dive.

* Specific issues under Education, analyse the account heads for budget expenditure and district spending trends
* Identify some key fund flow issues and/or interesting data insights/stories

#### **Read**

Read the cleaned dataset after processing through the Data Health Report

```{r read, cache = TRUE}
# State expenditure data
budget_exp_final <- read_csv("../data/processed/budget_exp_final.csv")

# Distirct spending data
district_spending_exp_final <- read_csv("../data/processed/district_spending_exp_final.csv")
```

#### **Clean**

Normalize the data fields across files for analysis.

```{r clean, cache = TRUE}
# Format the monetary value fields
# State Budget Expenditure
budget_exp_final$sanction <- accounting(budget_exp_final$sanction)
budget_exp_final$addition <- accounting(budget_exp_final$addition)
budget_exp_final$savings <- accounting(budget_exp_final$savings)
budget_exp_final$revised <- accounting(budget_exp_final$revised)

# District Spending  Expenditure
district_spending_exp_final$gross <- accounting(district_spending_exp_final$gross)
district_spending_exp_final$agded <- accounting(district_spending_exp_final$agded)
district_spending_exp_final$btded <- accounting(district_spending_exp_final$btded)
district_spending_exp_final$net_payment <- accounting(district_spending_exp_final$net_payment)
```

### **Sector Analysis**

Analysis of Education Department entries in the FY 2018-19 budget expenditure datasets.

```{r overview-state, warning = FALSE}
# Account Heads
state_exp_ed_18_19 <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19") %>%
    group_by(major_desc, sub_major_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = (revised - sanction)/sanction) %>%
    arrange(desc(revised))

# Rename Fields
names(state_exp_ed_18_19) <- c("Major", "Sub Major", "Sanction", "Revised", "Percent")

# Format Fields
state_exp_ed_18_19$Percent <- formattable::percent(state_exp_ed_18_19$Percent)

# Print formatted table
formattable(state_exp_ed_18_19, list(
                                    Sanction = color_tile("coral", "tomato"),
                                    Revised = color_tile("lightgreen", "seagreen"),
                                    Percent = color_tile("paleturquoise", "turquoise")
                                    )
            )
```

```{r overview-dist-2, warning = FALSE}
# Account Heads
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19") %>%
    group_by(district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment)) %>%
    mutate(district = as_factor(district)) %>%
    ggplot(aes(x = district, y = net_payment, label = net_payment)) +
        geom_bar(stat = "identity") + scale_y_continuous(labels = comma) +
        geom_label(size = 3, hjust = 0.2) +
        labs(title = "District Wise Education Spending", x = "", y = "Amount") +
        theme_minimal() + theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
```

**Findings**

---

* ~98% of all state education expenditure was towards General Education function.
* The top sub-sectors were Elementary Education (53.76%), Secondary Education (38.54%) & University and Higher Education (7.24%).
* Lahaul and Spiti do not have any education spending in 2018-19; even with schools present in the area.
* **_Areas of Exploration_**:

  1. WELFARE OF SCHEDULED CASTES SC
  2. SOCIAL SECURITY AND WELFARE
  3. LANGUAGE DEVELOPMENT
  4. ART AND CULTURE

#### **Welfare Of Scheduled Castes**

Review on the welfare funding under state expenditure FY 2018-19.

```{r welfare-1}
# Welfare of Scheduled Castes SC
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "WELFARE OF SCHEDULED CASTES SC") %>%
    group_by(sub_minor_desc) %>%
    summarise(Sanction = sum(sanction), Revised = sum(revised)) %>%
    gather(type, amount, -sub_minor_desc) %>%
    ggplot(aes(x = type, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity") + scale_y_continuous(labels = comma) +
        geom_text(hjust = 0.5) +
        labs(title = "Welfare of Scheduled Castes Budget", x = "", y = "Amount") +
        theme_minimal() + coord_flip()
```

* Welfare Of Scheduled Castes SC had ₹36,200 lacs under sanctions, revised to zero later.
* The amount was allocated against Sub-Minor: SCHOLARSHIPS TO MINORITY COMMUNITIES [C00N]
* There is no district level spending under WELFARE OF SCHEDULED CASTES SC major head, confirming the findings in the state expenditure file.

```{r ed-across-sc, warning = FALSE}
# Education Spending
sc_state_exp_ed <- budget_exp_final %>%
    filter(year == "2018-19") %>%
    filter(demand_desc =="SCHEDULED CASTES SUB PLAN" & 
           stringr::str_detect(major_desc, pattern = "EDUC")) %>%
    group_by(demand_desc, major_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = (revised - sanction)/sanction) %>%
    arrange(desc(revised))
    
# Format % Field
sc_state_exp_ed$percent <- formattable::percent(sc_state_exp_ed$percent)

# Rename Fields
names(sc_state_exp_ed) <- c("Demand Desc", "Major Desc", "Sanction", "Revised", "Percent")

# Print formatted table
formattable(sc_state_exp_ed,
            list(
              Sanction = color_tile("coral", "tomato"),
              Revised = color_tile("lightgreen", "seagreen"),
              Percent = color_tile("paleturquoise", "turquoise")
              )
)
```

Post further exploration, it was discovered there is education spending under the SCHEDULED CASTES SUB PLAN.

#### **Social Security and Welfare**

```{r welfare-2}
# Social Security and Welfare
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "SOCIAL SECURITY AND WELFARE") %>%
    group_by(sub_minor_desc) %>%
    summarise(Sanction = sum(sanction), Revised = sum(revised)) %>%
    gather(type, amount, -sub_minor_desc) %>%
    ggplot(aes(x = type, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity") + scale_y_continuous(labels = comma) +
        geom_text(hjust = 0.5) +
        labs(title = "Social Security and Welfare", x = "", y = "Amount") +
        theme_minimal() + coord_flip()
```

* Social Security and Welfare had a little over ₹9,000 lacs under state budget expenditure.
* The amount was allocated against Sub-Minor: WELFARE OF HANDICAPPED CHILDREN [C00N]; 0.004% of the total education state budget.

```{r welfare-dist}
# Social Security and Welfare
welfare_dist_exp <- district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "SOCIAL SECURITY AND WELFARE") %>%
    group_by(district, ddo_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment))

# Rename Fields
names(welfare_dist_exp) <- c("District", "DDO", "Net Payment")

# Print formatted table
formattable(welfare_dist_exp, 
            list(
                `Net Payment` = color_tile("coral", "tomato")
                )
            )
```

* Only 4 districts have Social Security and Welfare Spending, towards welfare of Handicapped Children.
* All the spending is towards the SOE of salaries.
* 5 DDOs from the 4 districts; which are Govt. Senior Secondary and Higher Secondary schools.

#### **Language Development**

```{r lang-dev}
# Language Development
lang_dev <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             sub_major_desc == "LANGUAGE DEVELOPMENT") %>%
    group_by(sub_minor_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = (revised - sanction)/sanction) %>%
    arrange(desc(revised))
    
# Format % Field
lang_dev$percent <- formattable::percent(lang_dev$percent)

# Rename Fields
names(lang_dev) <- c("Sub-Minor", "Sanction", "Revised", "Percent")

# Print formatted table
formattable(lang_dev,
            list(
              Sanction = color_tile("coral", "tomato"),
              Revised = color_tile("lightgreen", "seagreen"),
              Percent = color_tile("paleturquoise", "turquoise")
              )
)
```

* All the spending under language development is for Sanskrit Pathshalas.
* In 2019, Sanskrit was declared as the second official language of HP.

```{r lang-dev-dist}
# Language Development District/SOEs
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             sub_major_desc == "LANGUAGE DEVELOPMENT") %>%
    group_by(district, soe_description) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment)) %>%
    ggplot(aes(x = district, y = net_payment, fill = soe_description)) +
        geom_bar(stat = "identity", position = "stack") +
        scale_y_continuous(labels = comma) +
        labs(title = "Language Development Spending towards Sanskrit Pathshalas", 
             x = "", y = "Net Payment") +
        scale_fill_discrete(name = "SOE") + 
        theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

# Language Development
lang_dev_dist <- district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             sub_major_desc == "LANGUAGE DEVELOPMENT") %>%
    group_by(district, ddo_desc) %>%
    summarise(net_payment = sum(net_payment))

# Rename fields
names(lang_dev_dist) <- c("District", "DDO", "Net Payment")

# Print formatted table
formattable(lang_dev_dist, 
            list(
                `Net Payment` = color_tile("coral", "tomato")
                )
            )
```

* 5 districts are spending towards Language Development (Sanskrit Pathshalas); mostly on Salaries.
* Most DDOs are Sanskrit Colleges working in these district areas.

#### **Art and Culture**

```{r art}
# Art and culture
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "ART AND CULTURE") %>%
    group_by(soe_description) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    gather(type, amount, -soe_description) %>%
    mutate(type = as_factor(type)) %>%
    ggplot(aes(x = soe_description, y = amount, fill = type)) +
        geom_bar(stat = "identity", position = "dodge") +
        scale_y_continuous(labels = comma) + 
        scale_fill_discrete(name = "Type", labels = c("Sanction", "Revised")) + 
        labs(title = "Art and Culture State Budget Expenditure", x = "SOE", y = "Amount") +
        theme_minimal() + coord_flip()
```

All the Arts and Culture state budget expenditure was towards State and District Public Libraries, with the majority of amount going towards salaries.

```{r art-dist}
# Art and culture spending over time
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "ART AND CULTURE") %>%
    mutate(month = floor_date(transaction_date, unit = "month")) %>%
    group_by(month) %>%
    summarise(gross = sum(gross),
              net_payment = sum(net_payment)) %>%
    gather(type, amount, -month) %>%
    ggplot(aes(x = month, y = amount, color = type)) +
    geom_line() + geom_point() +
    scale_y_continuous(labels = comma) +
    scale_color_discrete(name = "Cylinders", labels = c("Gross", "Net Payment")) +
    labs(title = "Art and Culture District Spending Over Time", x = "Time", y = "Amount") +
    theme_minimal()

# Art and culture district spending over time
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "ART AND CULTURE") %>%
    mutate(month = floor_date(transaction_date, unit = "month")) %>%
    group_by(month, district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    ggplot(aes(x = month, y = net_payment, fill = district)) +
    geom_area(stat = "identity", position = "stack") +
        scale_y_continuous(labels = comma) +
        labs(title = "Arts and Culture Spending by Districts", 
             x = "", y = "Net Payment") +
        scale_fill_discrete(name = "District") + 
        theme_minimal()
```

* District level spending peaks in Jun, '18, Feb, '19 and Mar, '19
* Large spikes in spending in Shimla (Jun, '18) and Harmirpur (Mar, '19)

```{r art-dist-2}
# Art and culture district spending by DDO
art_ddo_dist_exp <- district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "ART AND CULTURE") %>%
    group_by(district, ddo_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    arrange(desc(net_payment)) %>%
    head(10)

# Rename fields
names(art_ddo_dist_exp) <- c("District", "DDO", "Net Payment")

# Print formatted table
formattable(art_ddo_dist_exp, 
            list(
                `Net Payment` = color_tile("coral", "tomato")
                )
            )
```

CHIEF LIB.C.S.LIBRARY SOLAN	; DDO with the largest spending expenditure on Arts and Culture, followed by A.D.(EDU.) STATE LIBRARY SHIMLA.

#### **Captial Outlay & Public Works**

```{r capital}
# Capital Outlay
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "CAPITAL OUTLAY  ON EDUCATION S") %>%
    group_by(minor_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    gather(type, amount, -minor_desc) %>%
    mutate(type = as_factor(type)) %>%
    ggplot(aes(x = minor_desc, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity", position = "dodge") +
        scale_y_continuous(labels = comma) + geom_text(position = position_dodge(width = 0.8)) +
        scale_fill_discrete(name = "Type", labels = c("Sanction", "Revised")) + 
        labs(title = "Capital Outlay on Education", x = "Minor", y = "Amount") +
        theme_minimal() + coord_flip()

# SOE
capital_outlay_ed <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "CAPITAL OUTLAY  ON EDUCATION S") %>%
    group_by(soe_description) %>%
    summarise(sanction = sum(sanction), revised = sum(revised))

# Rename fields
names(capital_outlay_ed) <- c("SOE", "Sanction", "Revised")

# Print formatted table
formattable(capital_outlay_ed, 
            list(
              Sanction = color_tile("coral", "tomato"),
              Revised = color_tile("lightgreen", "seagreen")                   
            )
)
```

* Capital Outlay on Education was highest for University and Higher Education (54%) and lowest for Elelmentary Education (13%)
* Over 99% of all Capital Outlay state expenditure was towards Major Works statement of expenditure.

```{r public-works}
# Public Works
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "PUBLIC WORKS") %>%
    group_by(sub_minor_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    gather(type, amount, -sub_minor_desc) %>%
    mutate(type = as_factor(type)) %>%
    ggplot(aes(x = sub_minor_desc, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity", position = "dodge") +
        scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
        scale_y_continuous(labels = comma) + 
        geom_text(position = position_dodge(width = 0.8)) +
        scale_fill_discrete(name = "Type", labels = c("Sanction", "Revised")) + 
        labs(title = "Public Works spending in Education", x = "Sub-Minor", y = "Amount") +
        theme_minimal() + coord_flip()
```

* The state expenditure under the Major Public Works, over 77% was towards - MAINTENANCE OF PRIMARY SCHOOL BUILDINGS. [S00N].

### **Additional Ideas**

#### **General Education**

Observation of various Minor account heads under General Education to understand trends.

```{r gen-ed}
# Minor
gen_ed_minor <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "GENERAL EDUCATION") %>%
    group_by(minor_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = formattable::percent(revised - sanction)/sanction) %>%
    arrange(desc(revised))

# Rename fields
names(gen_ed_minor) <- c("Minor", "Sanction", "Revised", "Percent")

# Print formatted table
formattable(gen_ed_minor, list(
                            Sanction = color_tile("coral", "tomato"),
                            Revised = color_tile("lightgreen", "seagreen"),
                            Percent = color_tile("paleturquoise", "turquoise")
                            )
            )
```

**Further Questions**

---

* Teacher Training as a share of Education under different levels.
* Composition of spending under Sarav Sikhsha Abhyan.
* Spending on assistance to non govt. educational organisations.

#### **Education Account Heads**

```{r ed-across, warning = FALSE}
# Education Spending
budget_exp_final %>%
    filter(year == "2018-19") %>%
    filter(stringr::str_detect(demand_desc, pattern = "EDUC") | 
           stringr::str_detect(major_desc, pattern = "EDUC")) %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    arrange(desc(sanction)) %>%
    gather(type, amount, -demand_desc) %>%
    mutate(demand_desc = as_factor(demand_desc)) %>%
    ggplot(aes(x = demand_desc, y = amount, fill = type)) +
        geom_bar(stat = "identity", position = position_dodge()) +
        scale_x_discrete(label = function(x) stringr::str_trunc(x, 12)) +
        scale_y_continuous(labels = comma) +
        labs(title = "Education Expenditure Across Demand Account Heads", x = "Demand", y = "Amount") +
        theme_minimal() + theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

Education specific account heads are across all demand heads.

**General Education Across Demand Account Heads**

```{r gen-ed-across, warning = FALSE}
# Education Spending
gen_education_across <- budget_exp_final %>%
    filter(year == "2018-19" & major_desc == "GENERAL EDUCATION") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = (revised - sanction)/sanction) %>%
    arrange(desc(revised))
    
# Format % Field
gen_education_across$percent <- formattable::percent(gen_education_across$percent)

# Rename Fields
names(gen_education_across) <- c("Demand", "Sanction", "Revised", "Percent")

# Print formatted table
formattable(gen_education_across,
            list(
              Sanction = color_tile("coral", "tomato"),
              Revised = color_tile("lightgreen", "seagreen"),
              Percent = color_tile("paleturquoise", "turquoise")
              )
)
```

* General Education expenditure is spread across various account heads, requireing further exploration.
* There is decreased is revised state budget expenditure across the different demand account heads.

#### **Major Account Heads Codes**

Begins with | Accounts 
--- | ---
20 | Administration
24/25/27 | Agricultural/Rural
30 | Transportation/Tourism
4 | Capital Outlay
5 | Capital Outlay towards Transporation/Tourism
6 | Debts and Loans
7 | Loans to Govt. Servants

### **Session Info**

```{r session}
# Information about the machine and enviroment
sessionInfo()
```
