---
title: '**Deep Dive Report**'
author: "TheDataAreClean"
date: "6th April, '20"
output:
  html_document:
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

# Load packages
library(tidyverse)
library(formattable)
library(lubridate)
library(scales)
library(ggalt)
library(treemapify)
library(ggplotify)
```

### **Objective**

Follow-up on initial assignment with a deep dive.

* Specific issues under Education, analyse the account heads for budget expenditure and district spending trends
* Identify some key fund flow issues and/or interesting data insights/stories

#### **Read**

Read the cleaned dataset after processing through the Data Health Report

```{r read, cache = TRUE}
# State expenditure data
budget_exp_final <- read_csv("../data/processed/budget_exp_final.csv")

# Distirct spending data
district_spending_exp_final <- read_csv("../data/processed/district_spending_exp_final.csv")
```

#### **Clean**

Normalize the data fields across files for analysis.

```{r clean, cache = TRUE}
# Format the monetary value fields
# State Budget Expenditure
budget_exp_final$sanction <- accounting(budget_exp_final$sanction)
budget_exp_final$addition <- accounting(budget_exp_final$addition)
budget_exp_final$savings <- accounting(budget_exp_final$savings)
budget_exp_final$revised <- accounting(budget_exp_final$revised)

# District Spending  Expenditure
district_spending_exp_final$gross <- accounting(district_spending_exp_final$gross)
district_spending_exp_final$agded <- accounting(district_spending_exp_final$agded)
district_spending_exp_final$btded <- accounting(district_spending_exp_final$btded)
district_spending_exp_final$net_payment <- accounting(district_spending_exp_final$net_payment)
```

### **Sector Analysis**

Analysis of Education Department entries in the budgetary datasets.

#### **State Budget Expenditure**

FY 2018-19 state budget expenditure transactions.

```{r overview, warning = FALSE}
# Account Heads
state_exp_ed_18_19 <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19") %>%
    group_by(major_desc, sub_major_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = (revised - sanction)/sanction) %>%
    arrange(desc(sanction))

# Rename Fields
names(state_exp_ed_18_19) <- c("Major", "Sub Major", "Sanction", "Revised", "Percent")

# Format Fields
state_exp_ed_18_19$Percent <- formattable::percent(state_exp_ed_18_19$Percent)

# Print formatted table
formattable(state_exp_ed_18_19, list(
                                    Sanction = color_tile("coral", "tomato"),
                                    Revised = color_tile("lightgreen", "seagreen"),
                                    Percent = color_tile("paleturquoise", "turquoise")
                                    )
            )
```

**Findings**

---

* ~98% of all state education expenditure was towards General Education function.
* The top sub-sectors were Elementary Education (53.76%), Secondary Education (38.54%) & University and Higher Education (7.24%).
* Adult Education, also part of General Education had ₹724 under sanctions, going down to zero under revised.

##### **Welfare**

Review on the welfare funding under state expenditure FY 2018-19.

```{r welfare-1}
# Welfare of Scheduled Castes SC
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "WELFARE OF SCHEDULED CASTES SC") %>%
    group_by(sub_minor_desc) %>%
    summarise(Sanction = sum(sanction), Revised = sum(revised)) %>%
    gather(type, amount, -sub_minor_desc) %>%
    ggplot(aes(x = type, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity") + scale_y_continuous(labels = comma) +
        geom_text(hjust = 0.5) +
        labs(title = "Welfare of Scheduled Castes Budget", x = "", y = "Amount") +
        theme_minimal() + coord_flip()
```

* Welfare Of Scheduled Castes SC had ₹36,200 under sanctions, revised to zero later.
* The amount was alloted against Sub-Minor: SCHOLARSHIPS TO MINORITY COMMUNITIES [C00N]

```{r welfare-2}
# Social Security and Welfare
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "SOCIAL SECURITY AND WELFARE") %>%
    group_by(sub_minor_desc) %>%
    summarise(Sanction = sum(sanction), Revised = sum(revised)) %>%
    gather(type, amount, -sub_minor_desc) %>%
    ggplot(aes(x = type, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity") + scale_y_continuous(labels = comma) +
        geom_text(hjust = 0.5) +
        labs(title = "Social Security and Welfare", x = "", y = "Amount") +
        theme_minimal() + coord_flip()
```

* Social Security and Welfare had a little over ₹9,000 under state budget expenditure.
* The amount was alloted against Sub-Minor: WELFARE OF HANDICAPPED CHILDREN [C00N]; 0.004% of the total education state budget.

##### **Language Development**

```{r lang-dev}
# Language Development
lang_dev <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             sub_major_desc == "LANGUAGE DEVELOPMENT") %>%
    group_by(sub_minor_desc, soe_description) %>%
    summarise(amount = sum(revised)) %>%
    arrange(sub_minor_desc, desc(amount))

# Rename Fields
names(lang_dev) <- c("Sub-Minor", "SOE", "Revised")

# Print formatted table
formattable(lang_dev, 
            list(Revised = color_tile("paleturquoise", "turquoise"))
            )
```

* All the spending under language development is for Sanskrit Pathshalas.
* In 2019, Sanskrit was declared as the second official language of HP.

##### **Art and Culture**

```{r art}
# Art and culture
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "ART AND CULTURE") %>%
    group_by(soe_description) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    gather(type, amount, -soe_description) %>%
    mutate(type = as_factor(type)) %>%
    ggplot(aes(x = soe_description, y = amount, fill = type)) +
        geom_bar(stat = "identity", position = "dodge") +
        scale_y_continuous(labels = comma) + 
        scale_fill_discrete(name = "Type", labels = c("Sanction", "Revised")) + 
        labs(title = "Art and Culture State Budget Expenditure", x = "SOE", y = "Amount") +
        theme_minimal() + coord_flip()
```

All the Arts and Culture state budget expenditure was towards State and District Public Libraries, with the majority of amount going towards salaries.

##### **Captial Outlay & Public Works**

```{r capital}
# Capital Outlay
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "CAPITAL OUTLAY  ON EDUCATION S") %>%
    group_by(minor_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    gather(type, amount, -minor_desc) %>%
    mutate(type = as_factor(type)) %>%
    ggplot(aes(x = minor_desc, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity", position = "dodge") +
        scale_y_continuous(labels = comma) + geom_text(position = position_dodge(width = 0.8)) +
        scale_fill_discrete(name = "Type", labels = c("Sanction", "Revised")) + 
        labs(title = "Capital Outlay on Education", x = "Minor", y = "Amount") +
        theme_minimal() + coord_flip()

# SOE
capital_outlay_ed <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "CAPITAL OUTLAY  ON EDUCATION S") %>%
    group_by(soe_description) %>%
    summarise(sanction = sum(sanction), revised = sum(revised))

# Rename fields
names(capital_outlay_ed) <- c("SOE", "Sanction", "Revised")

# Print formatted table
formattable(capital_outlay_ed, 
            list(
              Sanction = color_tile("coral", "tomato"),
              Revised = color_tile("lightgreen", "seagreen")                   
            )
)
```

* Capital Outlay on Education was highest for University and Higher Education (54%) and lowest for Elelmentary Education (13%)
* Over 99% of all Capital Outlay state expenditure was towards Major Works statement of expenditure.

```{r public-works}
# Public Works
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "PUBLIC WORKS") %>%
    group_by(sub_minor_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    gather(type, amount, -sub_minor_desc) %>%
    mutate(type = as_factor(type)) %>%
    ggplot(aes(x = sub_minor_desc, y = amount, fill = type, label = amount)) +
        geom_bar(stat = "identity", position = "dodge") +
        scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
        scale_y_continuous(labels = comma) + 
        geom_text(position = position_dodge(width = 0.8)) +
        scale_fill_discrete(name = "Type", labels = c("Sanction", "Revised")) + 
        labs(title = "Public Works spending in Education", x = "Sub-Minor", y = "Amount") +
        theme_minimal() + coord_flip()
```

* The state expenditure under the Major Public Works, over 77% was towards - MAINTENANCE OF PRIMARY SCHOOL BUILDINGS. [S00N].

##### **General Education**

```{r gen-ed}
# Sub-Major
gen_ed_sub_major <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "GENERAL EDUCATION") %>%
    group_by(sub_major_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = formattable::percent(revised - sanction)/sanction) %>%
    arrange(desc(revised))

# Rename fields
names(gen_ed_sub_major) <- c("Sub-Major", "Sanction", "Revised", "Percent")

# Print formatted table
formattable(gen_ed_sub_major, list(
                                Sanction = color_tile("coral", "tomato"),
                                Revised = color_tile("lightgreen", "seagreen"),
                                Percent = color_tile("paleturquoise", "turquoise")
                                )
            )

# Minor
gen_ed_minor <- budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & year == "2018-19" &
             major_desc == "GENERAL EDUCATION") %>%
    group_by(minor_desc) %>%
    summarise(sanction = sum(sanction), revised = sum(revised)) %>%
    mutate(percent = formattable::percent(revised - sanction)/sanction) %>%
    arrange(desc(revised))

# Rename fields
names(gen_ed_minor) <- c("Minor", "Sanction", "Revised", "Percent")

# Print formatted table
formattable(gen_ed_minor, list(
                            Sanction = color_tile("coral", "tomato"),
                            Revised = color_tile("lightgreen", "seagreen"),
                            Percent = color_tile("paleturquoise", "turquoise")
                            )
            )
```

### **Session Info**

```{r session}
# Information about the machine and enviroment
sessionInfo()
```
