---
title: "**Exploratory Analysis Report**"
author: "TheDataAreClean"
date: "25th March, '20"
output: 
  html_document: 
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 6
---

### **Objective**

Exploratory analysis of the data to unsterstand the composition and trends of the budget expenditure and district spending data.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

# Load packages
library(tidyverse)
library(formattable)
library(lubridate)
```

### **Read**

Read the combined and cleaned state expenditure and district spending datasets for exploratory analysis.

```{r read, cache = TRUE}
# State expenditure data
budget_exp_final <- read_csv("../data/processed/budget_exp_final.csv")

# Distirct spending data
district_spending_exp_final <- read_csv("../data/processed/district_spending_exp_final.csv")
```

### **Analysis**

#### **State Budget Expenditure**

##### **Overall**

```{r overall}
# Quick Summary
budget_exp_final %>%
    group_by(year) %>%
    summarise(sanction = sum(sanction),
              addition = sum(addition),
              savings = sum(savings),
              revised = sum(revised))
```

* The sanctioned budget amount for Himachal Pradesh in 2018-19 was 159.7 Cr., a 14.5% increase over the previous years budget. 
* Post revisions, the allocated amount went down to 97.2% of the sanctions allocated in the year 2017-18, and 95% in the year 2018-19.

##### **Top Major Account Heads**

```{r account-heads}
# Account Head Wise spending
budget_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(demand_desc, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)
```

* Finance was alloyed the highest amount of all account heads, 30.49% of total sanctioned expenditure for the year; followed by Education with 14.12% & Public Works – Roads Bridges & Buildings at 10.26%
* This trend has followed form 2017-18 in terms of priority for the state budget allocations.

##### **Allocations Over Time**

```{r over-time}
# Trends for budget allocation over time
budget_exp_final %>%
    mutate(month = floor_date(transaction_date, unit = "month")) %>%
    group_by(month) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    gather(type, amount, -month) %>%
    ggplot(aes(x = month, y = amount, color = type)) +
    geom_line() + geom_point()
```

* Looking at the time trends data, the impression recieved is that budgets and revisions are done with the intent to distribute the aloocations at a dataily basis on the projected price.
* It will be interesting to compare it against the spending data to see the spending trends on ground.

##### **Voted/Charged & Plan/Non-Plan**

```{r voted-charged-plan, warning = FALSE}
# Voted/Charged
budget_exp_final %>%
    group_by(year, voted_charged) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(voted_charged, sanction) %>%
    mutate(percent_charged = percent(C/(V+C)),
           percent_voted = percent(V/(V+C)))

# Plan/Non-Plan
budget_exp_final %>%
    group_by(year, plan_nonplan) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(plan_nonplan, sanction) %>%
    mutate(percent_non_plan = percent(N/(N+P)),
           percent_plan = percent(P/(N+P)))
```

* ~83% of all budgetary allocation in Himachal Pradesh in Voted, remaining ~17% charged.
* ~78% of all budgetary allocation is non-plan expense, and remaning ~22% planned.

##### **2018-19 vs 2017-18**

```{r account-heads-exp}
# Overall Top 10 Investments
# Absolute change in account heads expenditure
budget_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
          percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    arrange(desc(delta))

# % change in account heads expenditure
budget_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
          percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    arrange(desc(percent_delta))

# Overall Top 10 Disinvestments
# Absolute change in account heads expenditure
budget_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
          percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    arrange(delta)

# Absolute change in account heads expenditure
budget_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
          percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    arrange(percent_delta)
```

* Finance, Education & Public Works has seen the most increase in absolute sanctions numbers in 2018-19 over 2017-18.
* Tourism and Civil Aviation (61.84%), Industries Mineral Supplies and Information Technology (52.77%) and Agriculture (34.04%) have seen highest sanctions in terms of % change.
* Labour Employemnt and Training, Election, Governor of Council and Ministries has seen the most decrease in absolute sanctions numbers in 2018-19 over 2017-18.
* Election (37.58%), Labour Employemnt and Training (23.19%) have seen highest descrease sanctions in terms of % change.

##### **Estimates vs Revisions**

```{r revised-exp}
## Estimates to Revision
## 2017-18 positive absolute change
budget_exp_final %>%
    filter(year == "2017-18") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(desc(revision_delta))
    
## 2017-18 positive percent change
budget_exp_final %>%
    filter(year == "2017-18") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(desc(percent_delta))

## 2017-18 negative absolute change
budget_exp_final %>%
    filter(year == "2017-18") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(revision_delta)
    
## 2017-18 negative percent change
budget_exp_final %>%
    filter(year == "2017-18") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(percent_delta)

## 2018-19 positive absolute change
budget_exp_final %>%
    filter(year == "2018-19") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(desc(revision_delta))
    
## 2018-19 positive percent change
budget_exp_final %>%
    filter(year == "2018-19") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(desc(percent_delta))

## 2018-19 negative absolute change
budget_exp_final %>%
    filter(year == "2018-19") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(revision_delta)
    
## 2018-19 negative percent change
budget_exp_final %>%
    filter(year == "2018-19") %>%
    group_by(demand_desc) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised)) %>%
    mutate(revision_delta = (revised - sanction),
           percent_delta =  percent((revised - sanction)/sanction)) %>%
    arrange(percent_delta)
```

* Irrigation Water Supply and Sanitation & Finance saw one of the highest increase from sanctions to revisons in both 2017-18 & 2018-19.
* Both financial years saw over huge increases in election expenditure from sanctions to revised, 57.29% in 2017-18 and 74.37% in 2018-19
* Education, Scheduled Castes Sub Plan, Rural Developement, Tribal Developement saw one of the highest decrease from sanctions to revisons in both 2017-18 & 2018-19.

#### **District Spending**

##### **Overall**

```{r district-overall}
# Quick Summary
district_spending_exp_final %>%
    group_by(year) %>%
    summarise(gross = sum(gross),
              adged = sum(agded),
              btded = sum(btded),
              net_payment = sum(net_payment))
```

* The net payments amount for Himachal Pradesh in 2018-19 was 293.6 Bn., a 12.7% increase over the previous years budget. 

##### **Allocations Over Time**

```{r district-over-time}
# Trends for budget allocation over time
district_spending_exp_final %>%
    mutate(month = floor_date(transaction_date, unit = "month")) %>%
    group_by(month) %>%
    summarise(gross = sum(gross),
              net_payment = sum(net_payment)) %>%
    gather(type, amount, -month) %>%
    ggplot(aes(x = month, y = amount, color = type)) +
    geom_line() + geom_point()
```

* Majority of payments happend during the month of March, a trend that can traced to both years.

##### **Top Districts & Treasuries**

```{r distirct-treasuries}
# District Wise spending
district_spending_exp_final %>%
    group_by(year, district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`net_payment (2018-19)` = sum(`2018-19`),
           `net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`net_payment (2017-18)`)) %>%
    select(district, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# Treasury Wise spending
district_spending_exp_final %>%
    group_by(year, treasury) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`net_payment (2018-19)` = sum(`2018-19`),
           `net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`net_payment (2017-18)`)) %>%
    select(treasury, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)
```

* Shimla Capital Treasury & Shimla Distrcits combined cover 51.34% spending in Himachal Pradesh in 2018-19; and 50.8% in 2017-18
* Followed by Dharamsala to at 10.78% of total district spending in 2018-19 and 11.25% in 2017-18.

##### **Top Major Account Heads**

```{r district-account-heads}
# Account Head wise spending
district_spending_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`net_payment (2018-19)` = sum(`2018-19`),
           `net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`net_payment (2017-18)`)) %>%
    select(demand_desc, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# Demand Description is NA
district_spending_exp_final %>%
    filter(is.na(demand_desc)) %>%
    group_by(year, district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`net_payment (2018-19)` = sum(`2018-19`),
           `net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`net_payment (2017-18)`)) %>%
    select(district, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)
```

* Finance (16.04%), Education (11.49%) and Health and Family Welfare have the highest amount of net payments across both financial years.
* 26.39% of net payments have no demand description against them, when looking at the district data they are spread acros all with Shimla at top with 24.06% of the NA spend and Dharamsala at 14.91%.

##### **Change Over Years**

```{r district-account-heads-change}
# Abslolute Positive Change in Net Payments
district_spending_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
           percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    select(demand_desc, `2018-19`, `2017-18`, delta, percent_delta) %>%
    arrange(desc(delta))

# % Positive Change in Net Payments
district_spending_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
           percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    select(demand_desc, `2018-19`, `2017-18`, delta, percent_delta) %>%
    arrange(desc(percent_delta))

# Abslolute Negative Change in Net Payments
district_spending_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
           percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    select(demand_desc, `2018-19`, `2017-18`, delta, percent_delta) %>%
    arrange(delta)

# % Negative Change in Net Payments
district_spending_exp_final %>%
    group_by(year, demand_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    mutate(delta = `2018-19` - `2017-18`,
           percent_delta = percent((`2018-19` - `2017-18`)/`2017-18`)) %>%
    select(demand_desc, `2018-19`, `2017-18`, delta, percent_delta) %>%
    arrange(percent_delta)
```

* Power Development, Land Revenue and District Administration, Rural Development had the highest absokute increase in net payments YoY.
* Power Development (160.48%), LUrban Development Town and Country Planning and Housing (77.97%) highest % increase in net payments YoY.
* Planning and Backward Area Sub-Plan had the highest absolute decrease in net payments YoY.

#### **Education Sector Analysis**

##### **State Budget Expenditure**

###### **Overall**
```{r ed-overall}
budget_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year) %>%
    summarise(sanction = sum(sanction),
              revised = sum(revised))
```

* In 2017-18, the revised education budget was 95% of originally sanctioned.
* In 2017-18, the revised education budget was down to 85% of originally sanctioned.

###### **Account Heads**

```{r ed-account-heads}
# Major Expenditure 
budget_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, major_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(major_desc, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)

# Major Expenditure
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & major_desc == "GENERAL EDUCATION") %>%
    group_by(year, sub_major_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(sub_major_desc, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)

# Sub-major Expenditure
# Elementary Education
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "ELEMENTARY EDUCATION") %>%
    group_by(year, minor_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(minor_desc, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)

# Secondary Education
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "SECONDARY EDUCATION") %>%
    group_by(year, minor_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(minor_desc, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)

# Minor Expenditure
# Elementary Education
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "ELEMENTARY EDUCATION" &
             minor_desc == "GOVERNMENT PRIMARY SCHOOLS") %>%
    group_by(year, sub_minor_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(sub_minor_desc, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)

# Secondary Education
budget_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "SECONDARY EDUCATION" &
             minor_desc == "GOVERNMENT SECONDARY SCHOOLS") %>%
    group_by(year, sub_minor_desc) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(sub_minor_desc, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)
```

* ~98% of all state education expenditure was towards General Education account head.
* The top Sub Majors were Elementary Education (53.76%), Secondary Education (38.54%) & University and Higher Education (7.24%).
* Under Elementary Education, 85% of the expenditure  was towards Govt. Primary Schools & for Secondary Education, 98% of expenditure was sanctioned towards Govt. Secondary Schools

###### **Statement of Expenditure**

```{r ed-soe}
# Education Overall
budget_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, soe_description) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(year, sanction, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total Sanction (2018-19)` = sum(`2018-19`),
           `Total Sanction (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of Expenditure (2018-19)` = percent(`2018-19`/`Total Sanction (2018-19)`),
           `% of Expenditure (2017-18)` = percent(`2017-18`/`Total Sanction (2017-18)`)) %>%
    select(soe_description, `2018-19`, `% of Expenditure (2018-19)`, `2017-18`, `% of Expenditure (2017-18)`)
```

* ~82% of all education expenses were towards salaries.

###### **Voted/Charged & Plan/Non-Plan**

```{r ed-voted-charged-plan, warning = FALSE}
# Voted/Charged
budget_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, voted_charged) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(voted_charged, sanction) %>%
    mutate(percent_charged = percent(C/(V+C)),
           percent_voted = percent(V/(V+C)))

# Plan/Non-Plan
budget_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, plan_nonplan) %>%
    summarise(sanction = sum(sanction)) %>%
    spread(plan_nonplan, sanction) %>%
    mutate(percent_non_plan = percent(N/(N+P)),
           percent_plan = percent(P/(N+P)))
```

* All the Himachal Pradesh state expenditure towards education was voted for.
* ~16% of the expenditure was planned and the remaining unplanned.

##### **District Spending**

###### **Overall**

```{r ed-district-overall}
# Quick Summary
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year) %>%
    summarise(gross = sum(gross),
              net_payment = sum(net_payment))
```

* The net payments amount for Education in Himachal Pradesh in 2018-19 was 38 Bn., a 4% increase over the previous year. 

###### **Allocations Over Time**

```{r ed-district-over-time}
# Trends for budget allocation over time
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    mutate(month = floor_date(transaction_date, unit = "month")) %>%
    group_by(month) %>%
    summarise(gross = sum(gross),
              net_payment = sum(net_payment)) %>%
    gather(type, amount, -month) %>%
    ggplot(aes(x = month, y = amount, color = type)) +
    geom_line() + geom_point()
```

###### **Top Districts & Treasuries**

```{r ed-distirct-treasuries}
# District Wise spending
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`net_payment (2018-19)` = sum(`2018-19`),
           `net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`net_payment (2017-18)`)) %>%
    select(district, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# % of Every District's Spending
district_spending_exp_final %>%
    filter(year == "2018-19") %>%
    mutate(ed_flag = if_else(is.na(demand_desc) | demand_desc != "EDUCATION", 
                             "OTHER", "EDUCATION")) %>%
    group_by(ed_flag, district) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(ed_flag, net_payment, fill = 0) %>%
    mutate(percent_ed = percent(EDUCATION/(EDUCATION + OTHER))) %>%
    arrange(desc(EDUCATION))

# Treasury Wise spending
district_spending_exp_final %>%
    filter(year == "2018-19") %>%
    mutate(ed_flag = if_else(is.na(demand_desc) | demand_desc != "EDUCATION", 
                             "OTHER", "EDUCATION")) %>%
    group_by(ed_flag, treasury) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(ed_flag, net_payment, fill = 0) %>%
    mutate(percent_ed = percent(EDUCATION/(EDUCATION + OTHER))) %>%
    arrange(desc(EDUCATION))

# Treasuries with Highest % Share in Education
district_spending_exp_final %>%
    filter(year == "2018-19") %>%
    mutate(ed_flag = if_else(is.na(demand_desc) | demand_desc != "EDUCATION", 
                             "OTHER", "EDUCATION")) %>%
    group_by(ed_flag, treasury) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(ed_flag, net_payment, fill = 0) %>%
    mutate(percent_ed = percent(EDUCATION/(EDUCATION + OTHER))) %>%
    arrange(desc(percent_ed))
```

* Shimla Capital Treasury and Dharamsala have respectively 17.78% and 16.37% of all education spending expenditure.
* Treasuries like Kamrau, Kupvi and Bhalie have over 70% of their budget spending in education.

```{r ed-district-account-heads}
# Major Expenditure 
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, major_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(major_desc, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# Major Expenditure
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & major_desc == "GENERAL EDUCATION") %>%
    group_by(year, sub_major_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(sub_major_desc, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# Sub-major Expenditure
# Elementary Education
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "ELEMENTARY EDUCATION") %>%
    group_by(year, minor_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(minor_desc, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# Secondary Education
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "SECONDARY EDUCATION") %>%
    group_by(year, minor_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(minor_desc, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# Minor Expenditure
# Elementary Education
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "ELEMENTARY EDUCATION" &
             minor_desc == "GOVERNMENT PRIMARY SCHOOLS") %>%
    group_by(year, sub_minor_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(sub_minor_desc, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)

# Secondary Education
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION" & 
             major_desc == "GENERAL EDUCATION" &
             sub_major_desc == "SECONDARY EDUCATION" &
             minor_desc == "GOVERNMENT SECONDARY SCHOOLS") %>%
    group_by(year, sub_minor_desc) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(sub_minor_desc, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)
```

* 96.42% of all net payments in education were towards General Education
* General Education net payments consists of Elementary (54.28%) & Secondary (36.37%)

###### **Statement of Expenditure**

```{r ed-district-soe}
# Education Overall
district_spending_exp_final %>%
    filter(demand_desc == "EDUCATION") %>%
    group_by(year, soe_description) %>%
    summarise(net_payment = sum(net_payment)) %>%
    spread(year, net_payment, fill = 0) %>%
    arrange(desc(`2018-19`)) %>%
    mutate(`Total net_payment (2018-19)` = sum(`2018-19`),
           `Total net_payment (2017-18)` = sum(`2017-18`)) %>%
    mutate(`% of net_payment (2018-19)` = percent(`2018-19`/`Total net_payment (2018-19)`),
           `% of net_payment (2017-18)` = percent(`2017-18`/`Total net_payment (2017-18)`)) %>%
    select(soe_description, `2018-19`, `% of net_payment (2018-19)`, `2017-18`, `% of net_payment (2017-18)`)
```

~78% of District spending expenditure is towards direct salries.

### **Session Info**

```{r session}
# Information about the machine and enviroment
sessionInfo()
```
